```{r  echo=FALSE, warning=FALSE, message=FALSE, results='asis'}
  metric_table <- data[['omim_table_metrics_new.txt']]

  kegg_metric_table <- read.table("../results/kegg/omim_kegg_comention_table_metrics.txt", sep ="\t", header = TRUE, stringsAsFactors = FALSE, quote = "")
  go_metric_table <- read.table("../results/go/omim_go_comention_table_metrics.txt", sep ="\t", header = TRUE, stringsAsFactors = FALSE, quote = "")
  reactome_metric_table <- read.table("../results/reactome/omim_reactome_comention_table_metrics.txt", sep ="\t", header = TRUE, stringsAsFactors = FALSE, quote = "")


  
  library(dplyr)
  library(kableExtra)
  library(knitr)
  library(ggplot2)
  library(igraph)


group_cluster_information <- function(df1, df2){
  clusters <- unique(df1$Cluster)
  cluster_list <- list()

  for (i in clusters){
    groups <- list()
    hp_groups <- filter(df1, Cluster == clusters[i])

    if (nrow(hp_groups) != 0) {
      groups[["hpo"]] <- hp_groups
    }
    system_groups <- filter(df2, Cluster == clusters[i])
    
    if (nrow(system_groups) != 0) {
      system_groups_sorted <- system_groups[order(-system_groups$Phenotype_coverage),]
      groups[["system"]] <- system_groups_sorted
      scores <- system_groups$Phenotype_coverage
            score <- max(scores)
            groups[["coherence_score"]] <- score
        }else{
            groups[["coherence_score"]] <- 0
        }
    name <- paste0("Cluster_*", i)
        cluster_list[[name]] <- groups 
  }


  return(cluster_list)
}

order_a_list_by_vector <- function(lst, field){
  length_vector <- c()

  for (i in 1:length(lst)){
    sublst <- lst[[i]]
    value <- sublst[[field]]
    length_vector[i] <- value
  } 
  ordered_lst <- lst[order(length_vector, decreasing=TRUE)] 

  return(ordered_lst)
}


produce_barplot <- function(column, title, ylab){
  data_frame <- summarySE(metric_table, measurevar=column, groupvars=c("Type"))
    
  plt <- ggplot(data_frame, aes(x=Type, y=get(column), fill=Type)) 
  plt <- plt + geom_bar(stat="identity")
  plt <- plt + geom_errorbar(aes(ymin=get(column)-sd, ymax=get(column)+sd),
                  width=.2,                    # Width of the error bars
                  position=position_dodge(.9)) #+
  plt <- plt + theme(axis.text.x = element_text(angle = 45, hjust = 1), 
                                axis.title.x=element_blank()
                                )
  plt <- plt + ylab(ylab)
  if(title != ''){
    plt <- plt + ggtitle(title) 
  }
  plt <- plt + theme(legend.position="none")
    
  return(plt)
}

produce_barplot_comention_kegg <- function(column, title, ylab){
  data_frame <- summarySE(kegg_metric_table, measurevar=column, groupvars=c("Type"))
    
  plt <- ggplot(data_frame, aes(x=Type, y=get(column), fill=Type)) 
  plt <- plt + geom_bar(stat="identity")
  plt <- plt + geom_errorbar(aes(ymin=get(column)-sd, ymax=get(column)+sd),
                  width=.2,                    # Width of the error bars
                  position=position_dodge(.9)) #+
  plt <- plt + theme(axis.text.x = element_text(angle = 45, hjust = 1), 
                                axis.title.x=element_blank()
                                )
  plt <- plt + ylab(ylab)
  if(title != ''){
    plt <- plt + ggtitle(title) 
  }
  plt <- plt + theme(legend.position="none")
    
  return(plt)
}

produce_barplot_comention_go <- function(column, title, ylab){
  data_frame <- summarySE(go_metric_table, measurevar=column, groupvars=c("Type"))
    
  plt <- ggplot(data_frame, aes(x=Type, y=get(column), fill=Type)) 
  plt <- plt + geom_bar(stat="identity")
  plt <- plt + geom_errorbar(aes(ymin=get(column)-sd, ymax=get(column)+sd),
                  width=.2,                    # Width of the error bars
                  position=position_dodge(.9)) #+
  plt <- plt + theme(axis.text.x = element_text(angle = 45, hjust = 1), 
                                axis.title.x=element_blank()
                                )
  plt <- plt + ylab(ylab)
  if(title != ''){
    plt <- plt + ggtitle(title) 
  }
  plt <- plt + theme(legend.position="none")
    
  return(plt)
}
produce_barplot_comention_reactome <- function(column, title, ylab){
  data_frame <- summarySE(reactome_metric_table, measurevar=column, groupvars=c("Type"))
    
  plt <- ggplot(data_frame, aes(x=Type, y=get(column), fill=Type)) 
  plt <- plt + geom_bar(stat="identity")
  plt <- plt + geom_errorbar(aes(ymin=get(column)-sd, ymax=get(column)+sd),
                  width=.2,                    # Width of the error bars
                  position=position_dodge(.9)) #+
  plt <- plt + theme(axis.text.x = element_text(angle = 45, hjust = 1), 
                                axis.title.x=element_blank()
                                )
  plt <- plt + ylab(ylab)
  if(title != ''){
    plt <- plt + ggtitle(title) 
  }
  plt <- plt + theme(legend.position="none")
    
  return(plt)
}

delete_parental_terms<- function(df){
  library(GOSim)
  library(GO.db)
  library(dplyr)
    
    
  setOntology(ont = "BP")
  go_l <- getParents()
    
  initial_clusters <- length(unique(df$Cluster))
  
  clusters_l <- unique(df$Cluster)
  df1 <- data.frame()
  
  for(cluster in clusters_l){
    cluster_subset <- filter(df, Cluster == cluster)
    funsys_l <- cluster_subset$"Cell function"
    
    parents_l <- c()
    if(length(funsys_l) > 1){
      for(funsys in funsys_l){
        parents_l <- c(parents_l, go_l[[funsys]])
      }
      for(parent in parents_l){
        if(parent %in% funsys_l){
          cluster_subset <- (cluster_subset[!cluster_subset$"Cell function" == parent, ])
        }
      }
      df1 <- rbind(df1, cluster_subset)
    }else{
      df1 <- rbind(df1, cluster_subset)
    }
    
  }
  return(df1)
}

load_comention_files <- function(data, column_names, path_column, header, col_name_vector, group1, group2, group3){
        factor_columns <- match(column_names, names(data))
        factor_combinations <- unique(data[column_names])
        all_data <- data.frame()
        for(row in 1:nrow(factor_combinations)){
                combination <- as.vector(t((factor_combinations[row,]))) #extract row AND convert to vector
                check_combination <- data[factor_columns] == combination[col(data[factor_columns])]
                name_list <- paste(combination, collapse='_')
                paths <- data[[path_column]][which(apply(check_combination, 1, sum) == length(combination))]
                count = 1

                for(file_path in paths){
                        files <- read.table(file_path, sep="\t", header=header)
                        files$V4 <- NULL
                        files <- cbind(files ,rep(c(paste(name_list)), nrow(files)))
                        files <- cbind(files ,rep(c(paste(name_list,count, sep="_")), nrow(files)))
                        if(startsWith(name_list, group1)){
                          files <- cbind(files ,rep(c(paste(group1)), nrow(files)))
                        }else if(startsWith(name_list, group2)){
                          files <- cbind(files ,rep(c(paste(group2)), nrow(files)))
                        }else{
                          files <- cbind(files ,rep(c(paste(group3)), nrow(files)))
                        }
                        colnames(files) <- col_name_vector
                        count <- count + 1

                        all_data <- rbind(all_data, files)   
                }             
        }
       
        return(all_data)

      }


produce_density_plot <- function(data_table, values, net_names, net_types, x_scale = FALSE, x_lab, title, facet = FALSE){
        plt <- ggplot(data_table, aes(x=get(values))) + geom_density(aes(group=get(net_names), colour=get(net_types)))
        col <- c("firebrick1", "deepskyblue2")

        if (facet == TRUE){
          plt <- plt + facet_wrap( ~ groups, scales = "free")
        }
        if(x_scale == TRUE){
          plt <- plt + scale_x_continuous(limits = c(0, 1))
        }
        plt <- plt + ylim(0, 4.5)
        plt <- plt + xlab(x_lab)
        plt <- plt + ylab("Density")
        plt <- plt + ggtitle(title)
        plt <- plt + theme(legend.title=element_blank())
        plt <- plt + scale_fill_manual(values = col)
        plt <- plt + scale_color_manual(values = col)

        return(plt)
      }


```
# OMIM Neuromuscular Diseases Study

 

## Step one: Datasets of NMDs, genes, phenotypes and typical NMD phenotypes  
  

OMIM offer a brief description of diseases.  
By conducting a literature review, we can stabilize a number of keywords to filter the OMIM diseases and get a certain number of neuromuscular diseases:  
- MUSCULAR DYSTROPHY.   
- MYOPATHY.   
- MYASTHENIC.   
- MYASTHENIA.   
- NEUROPATHY.   
- AMYOTROPHIC LATERAL SCLEROSIS.   
- SPINAL MUSCULAR ATROPHY.   
- SPINAL AND BULBAR MUSCULAR ATROPHY.   
- MYOTONIA.   
- PERIODIC PARALYSIS.   
- MYOTONIC DYSTROPHY.   
- MITOCHONDRIAL CYTOPATHY.   
- NECROTIZING ENCEPHALOMYELOPATHY.   
- MITOCHONDRIAL DNA DEPLETION.   
  
Once we have the list of neuromuscular diseases, we count the frequency of phenotypes in neuromuscular and non neuromuscular diseases. With these data, we can perform a fisher test analysis to find those phenotypes which are more represented in neuromusculare diseases rather than in non neuromuscular diseases.
  
<div align="center">**Table 1**: Preprocesing data summary</div>  

```{r  echo=FALSE, warning=FALSE, message=FALSE}
  summary_metric_table <- read.table("../results/omim_summary", sep="\t", header = FALSE)
  colnames(summary_metric_table) <- c("Metric", "Value")
  kable(summary_metric_table)   %>%
  kable_classic(full_width = F, html_font = "Calibri", font_size = 16 ) %>% row_spec(0, bold = TRUE) %>% column_spec(1, bold = TRUE)
  
```
  
    
## Step 2: Bipartite network to obtain phenotype-phenotype associations
  
<div align="center">**Table 2:** Bipartite results general summary.</div>  

```{r  echo=FALSE, warning=FALSE, message=FALSE}
  pairs_metric_table <- read.table("../results/omim_pairs_table", sep="\t", header = FALSE)
  colnames(pairs_metric_table) <- c("Metric", "Value")
  kable(pairs_metric_table) %>%
  kable_classic(full_width = F, html_font = "Calibri", font_size = 16 ) %>% row_spec(0, bold = TRUE) %>% column_spec(1, bold = TRUE)

```
  
  
In the following plot we can see the number of pairs which form the subnetwork with a HyI >= 2. 
  
**Figure 1:** Number of phenotype pairs with a HyI >= 2.

```{r  echo=FALSE, warning=FALSE, message=FALSE}
produce_barplot("pairs_number", "Significant Pairs Number OMIM", "Number of pairs")
  
```
    
  
<div align="center">**Table 3:** Number of phenotype pairs with a HyI >= 2.</div>  
   
```{r  echo=FALSE, warning=FALSE, message=FALSE}
kable(ddply(metric_table, .(Type), summarize, "Pairs Number" = paste(round(mean(pairs_number),3), "\U00B1", round(sd(pairs_number),3)))) %>%
kable_classic(full_width = F, html_font = "Calibri", font_size = 16 ) %>% row_spec(0, bold = TRUE) %>% column_spec(1, bold = TRUE)
  
```
  

### Validating phenotype-phenotye associations by PubMed co-mention
  
  

**Figure 2** Comention pvalue density plot.  

```{r  echo=FALSE, warning=FALSE, message=FALSE}

comention_data <- load_comention_files(metric_table, 'Type', 'comention_analysis', FALSE, c("HPOA", "HPOB", "Pvalue", "Net_type", "Net_name", "groups"), "NMDs", "non_NMDs")

produce_density_plot(comention_data, "Pvalue", "Net_name", "Net_type", x_scale = FALSE, "P-value", "P-value distribution", facet = FALSE)

```   
  

  
<div align="center">**Table 4** Number of co-occurrent pairs.</div>  

```{r  echo=FALSE, warning=FALSE, message=FALSE}
kable(ddply(metric_table, .(Type), summarize, "Pairs Number" = paste(round(mean(number_non_coocurrent_pairs_),3), "\U00B1", round(sd(number_non_coocurrent_pairs_),3)))) %>% 
kable_classic(full_width = F, html_font = "Calibri", font_size = 16 ) %>% row_spec(0, bold = TRUE) %>% column_spec(1, bold = TRUE)
  
```

  
  
## Step 3: Tripartite network to obtain phenotype-function pairs
  
In this step we will get phenotype-genes pairs. Through a functional assignment of these genes we can get phenotype-function pairs.
  
<div align="center">**Table 5** Tripartite results general summary.</div>  
    
```{r  echo=FALSE, warning=FALSE, message=FALSE}
  gene_metric_table <- read.table("../results/omim_gene_table", sep="\t", header = FALSE)
  colnames(gene_metric_table) <- c("Metric", "Value")
  kable(gene_metric_table) %>%
  kable_classic(full_width = F, html_font = "Calibri", font_size = 16 ) %>% row_spec(0, bold = TRUE) %>% column_spec(1, bold = TRUE)

```
  

### Validating phenotype-function associations by PubMed co-mention
  
#### KEGG
  
**Figure 3** Significant phenotype-KEGG pairs  

```{r  echo=FALSE, warning=FALSE, message=FALSE}
  produce_barplot_comention_kegg("number_coocurrent_pairs_kegg", "KEGG", "Pares Significativos")

``` 
  
  
<div align="center">**Table 6** Significant phenotype-KEGG pairs.</div>  

```{r  echo=FALSE, warning=FALSE, message=FALSE}
  
kable(ddply(kegg_metric_table, .(Type), summarize, "Co-occurrent pairs" = paste(round(mean(number_coocurrent_pairs_kegg),3), "\U00B1" , round(sd(number_coocurrent_pairs_kegg),3)), "No co-occurrent pairs" = paste(round(mean(number_non_coocurrent_pairs_kegg),3), "\U00B1" , round(sd(number_non_coocurrent_pairs_kegg),3))), align = 'c') %>% 
kable_classic(full_width = T, html_font = "Calibri", font_size = 16) %>% row_spec(0, bold = TRUE) %>% column_spec(1, bold = TRUE)
  
``` 
  
  
  
#### GO

  
**Figure 4** Significant phenotype-GO pairs.  

```{r  echo=FALSE, warning=FALSE, message=FALSE}
  produce_barplot_comention_go("number_coocurrent_pairs_go", "GO", "Pares Significativos")

``` 

  
<div align="center">**Table 7** Significant phenotype-GO pairs.</div>  
  
```{r  echo=FALSE, warning=FALSE, message=FALSE}
  
kable(ddply(go_metric_table, .(Type), summarize, "Co-occurrent pairs" = paste(round(mean(number_coocurrent_pairs_go),3), "\U00B1" , round(sd(number_coocurrent_pairs_go),3)), "No co-occurrent pairs" = paste(round(mean(number_non_coocurrent_pairs_go),3), "\U00B1" , round(sd(number_non_coocurrent_pairs_go),3))), align = 'c') %>% 
kable_classic(full_width = T, html_font = "Calibri", font_size = 16) %>% row_spec(0, bold = TRUE) %>% column_spec(1, bold = TRUE)

``` 
  
  
  
##### Reactome
  
**Figure 5** Significant phenotype-Reactome pairs.  

```{r  echo=FALSE, warning=FALSE, message=FALSE}
produce_barplot_comention_reactome("number_coocurrent_pairs_reactome", "Reactome ", "Pares Significativos")

```
  
  
<div align="center">**Table 8** Significant phenotype-Reactome pairs.</div> 

```{r  echo=FALSE, warning=FALSE, message=FALSE}
  
kable(ddply(reactome_metric_table, .(Type), summarize, "Co-occurrent pairs" = paste(round(mean(number_coocurrent_pairs_reactome),3), "\U00B1" , round(sd(number_coocurrent_pairs_reactome),3)), "No co-occurrent pairs" = paste(round(mean(number_non_coocurrent_pairs_reactome),3), "\U00B1" , round(sd(number_non_coocurrent_pairs_reactome),3))), align = 'c') %>% 
kable_classic(full_width = T, html_font = "Calibri", font_size = 16) %>% row_spec(0, bold = TRUE) %>% column_spec(1, bold = TRUE)

``` 
  
  
## Step 4: Cluster Analysis
  
With linkcomm R package we obtain a number of clusters for each network.  In the following plots we show some results of the clustering process. By one hand we have measured the number of clusters obtained and the average number of nodes of each cluster.  
  
**Figure 6** Cluster analysis summary. 

```{r  cho=FALSE, warning=FALSE, message=FALSE}
grid.arrange(
  produce_barplot("Cluster_number", "Cluster Number OMIM", "Number of clusters"),
  produce_barplot("Average_Cluster_size", "Cluster Size OMIM", "Average Cluster size"),
  ncol = 2
)

``` 
  

  
<div align="center">**Table 9** Cluster analysis summary.</div>  

```{r  cho=FALSE, warning=FALSE, message=FALSE}

kable(ddply(metric_table, .(Type), summarize, "Cluster Number" = paste(round(mean(Cluster_number),3), "\U00B1" , round(sd(Cluster_number),3)), "Average Cluster size" = paste(round(mean(Average_Cluster_size),3), "\U00B1" , round(sd(Average_Cluster_size),3))), align = 'c') %>% 
kable_classic(full_width = T, html_font = "Calibri", font_size = 16) %>% row_spec(0, bold = TRUE) %>% column_spec(1, bold = TRUE)
  
```
   

### Functionally-coherent clusters of phenotypes in NMDs
  
**Figure 7** Functional coherent clusters. 

```{r  cho=FALSE, warning=FALSE, message=FALSE}
grid.arrange(
  produce_barplot("Cluster_with_70_coherence_0.05", "70 % Coherence", "Number of clusters"),
  produce_barplot("Cluster_with_50_coherence_0.05", "50 % Coherence", "Number of clusters"),
  ncol = 2
)

``` 
  
  
<div align="center">**Table 10** Functional coherent clusters summary.</div>
    
```{r  cho=FALSE, warning=FALSE, message=FALSE}
  
kable(ddply(metric_table, .(Type), summarize, "Cluster with 70 coherence_0.05" = paste(round(mean(Cluster_with_70_coherence_0.05),3), "\U00B1" , round(sd(Cluster_with_70_coherence_0.05),3)), "Cluster with 50 coherence_0.05" = paste(round(mean(Cluster_with_50_coherence_0.05),3), "\U00B1" , round(sd(Cluster_with_50_coherence_0.05),3))), align = 'c') %>% 
kable_classic(full_width = T, html_font = "Calibri", font_size = 16) %>% row_spec(0, bold = TRUE) %>% column_spec(1, bold = TRUE)
```
  


## Clusters Details

In this section we include OMIM neuromuscular cluster's  details.  

```{r  cho=FALSE, warning=FALSE, message=FALSE, results='asis'}


clusters_hpo <- read.table("~/PhenoClusters/analysed_networks/OMIM/NMDs/clustering.R_0000/clusters_file_with_names", sep ="\t", quote="", stringsAsFactors = FALSE, header = TRUE)
colnames(clusters_hpo) <- c("HPO", "Cluster", "Description")
if(nrow(clusters_hpo) > 0){
    for (i in 1:length(clusters_hpo$HPO)){
      url <- paste0("https://hpo.jax.org/app/browse/term/", clusters_hpo$HPO[i])
      clusters_hpo$HPO[i] <- text_spec(clusters_hpo$HPO[i], link = url)
    }
  }

kegg <- read.table("~/PhenoClusters/analysed_networks/OMIM/NMDs/clusters_single_funsys_merger.py_0000/clusters_with_70_coherence_systems_kegg_0.05", sep ="\t", quote="", stringsAsFactors = FALSE, header = FALSE)
colnames(kegg) <- c("Cluster", "Cell function", "Description", "Phenotype_coverage", "Genes")
if(nrow(kegg) > 0){
    for (i in 1:length(kegg$"Cell function")){
      url <- paste0("https://www.genome.jp/dbget-bin/www_bget?pathway+", kegg$"Cell function"[i])
      kegg$"Cell function"[i] <- text_spec(kegg$"Cell function"[i], link = url)
    }
  }

go_df <- read.table("~/PhenoClusters/analysed_networks/OMIM/NMDs/clusters_single_funsys_merger.py_0001/clusters_with_70_coherence_systems_go_0.05", sep ="\t", quote="", stringsAsFactors = FALSE, header = FALSE)
colnames(go_df) <- c("Cluster", "Cell function", "Description", "Phenotype_coverage", "Genes")
go <- delete_parental_terms(go_df)
if(nrow(go) > 0){  
    for (i in 1:length(go$"Cell function")){
      url <- paste0("http://amigo.geneontology.org/amigo/term/", go$"Cell function"[i])
      go$"Cell function"[i] <- text_spec(go$"Cell function"[i], link = url)
    }
  }

reactome <- read.table("~/PhenoClusters/analysed_networks/OMIM/NMDs/clusters_single_funsys_merger.py_0002/clusters_with_70_coherence_systems_reactome_0.05", sep ="\t", quote="", stringsAsFactors = FALSE, header = FALSE)
colnames(reactome) <- c("Cluster", "Cell function", "Description", "Phenotype_coverage", "Genes")
if(nrow(reactome) > 0){  
    for (i in 1:length(reactome$"Cell function")){
      url <- paste0("https://reactome.org/content/detail/", reactome$"Cell function"[i])
      reactome$"Cell function"[i] <- text_spec(reactome$"Cell function"[i], link = url)
    }
  }

omim_clusters_system <- rbind(kegg, go, reactome)

all_clusters_lst <- group_cluster_information(clusters_hpo, omim_clusters_system)

ordered_cluster_lst <- order_a_list_by_vector(all_clusters_lst, "coherence_score")


#clusters_file
clusters_file_cytoscape <- read.table("~/PhenoClusters/analysed_networks/OMIM/NMDs/clustering.R_0000/cytoscape_file", sep=" ", header=FALSE, quote="", stringsAsFactors = FALSE)
clusters_file <- clusters_file_cytoscape[,c(1,3,5)]
colnames(clusters_file) <- c("HPO-A", "HPO-B", "Cluster")


for(i in 1:length(ordered_cluster_lst)){
    list <- ordered_cluster_lst[[i]]
    cluster_name <- unlist(strsplit(names(ordered_cluster_lst[i]), "_"))
    cluster_number <- unlist(strsplit(names(ordered_cluster_lst[i]), "\\*"))[2]
    if(!is.null(list[["system"]])){
      cat("_ _ _", "\n", "_ _ _", "\n", "_ _ _", "\n")
      list["coherence_score"] <- NULL
      cat("# ", cluster_name, "\n", "\n")

      cluster <- subset(clusters_file, Cluster == cluster_number)
      hpo_names <- subset(clusters_hpo, Cluster == cluster_number)
      g <- graph.data.frame(cluster, directed = FALSE)
      plot(g, vertex.label=gsub(" ", "\n", hpo_names$Description), vertex.shape="none", vertex.label.cex=c(0.9), vertex.label.color="black")

      for(i in list){
        print(knitr::kable(i, format="markdown", row.name=FALSE))
        cat("\n")
      }
    }
  }

for(i in 1:length(ordered_cluster_lst)){
    list <- ordered_cluster_lst[[i]]
    cluster_name <- unlist(strsplit(names(ordered_cluster_lst[i]), "_"))
    cluster_number <- unlist(strsplit(names(ordered_cluster_lst[i]), "\\*"))[2]
    if(is.null(list[["system"]])){
      cat("_ _ _", "\n", "_ _ _", "\n", "_ _ _", "\n")
      list["coherence_score"] <- NULL
      cat("# ", cluster_name, "\n", "\n")

      cluster <- subset(clusters_file, Cluster == cluster_number)
      hpo_names <- subset(clusters_hpo, Cluster == cluster_number)
      g <- graph.data.frame(cluster, directed = FALSE)
      plot(g, vertex.label=gsub(" ", "\n", hpo_names$Description), vertex.shape="none", vertex.label.cex=c(0.9), vertex.label.color="black")

      for(i in list){
        print(knitr::kable(i, format="markdown", row.name=FALSE))
        cat("\n")
      }
    }
  }

```